Mechanized Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-01-19 paper-new/categorical-model.tex
update definition of abstract model

@@

\begin{definition}\label{def:step-1-model}
A \emph{step-1 intensional model} consists of all the data of a step-0 intensional model along
with the additional categories and functors for bisimiarity described above.
\end{definition}

@@

@@

Now we can discuss what it means for an intensional model to model the dynamic type.

\begin{definition}\label{def:step-4-model}
  A \emph{step-4 intensional model}, or simply an \emph{intensional model}, consists
  of all the data of a step-3 intensional model, along with 
  %a distinguished value object $D \in \ob(\vf)$ such that:
  %
  \begin{enumerate}
    \item A distinguished value object $D \in \ob(\vf)$.
    (By definition of a step-3 model, this means there is also a monoid $\pv_D$ of
    perturbations and a homomorphism $\ptbv_D$.)
    \item For each value type $A$, there is a value relation $\text{inj}_A : A \rel D$.
    (By definition of a step-3 model, this relation satisfies the push-pull property and is
    quasi-left-representable.)
    \item If $c : A \rel A'$, then $\text{inj}_{A} = c \comp \text{inj}_{A'}$.
  \end{enumerate}
\end{definition}

@@

In the previous section, we have given the definition of an intensional model
of gradual typing as a series of steps with each definition building on the previous one.
%
Here, we discuss how to construct an extensional model from an intensional model.
We do so in several phases, beginning with a step-1 intensional model
and ending with an extensional model.
Moreover, this construction is \emph{modular}, in that each phase of the
construction does not depend on the details of the previous ones.
However, this process cannot proceed in isolation: some phases require
additional inputs. We will make clear what data must be supplied to each phase.

% Going from a step-1 intensional model $\mathcal{M}_1$ to a step-2 intensional model
% requires defining a bisimilarity relation on the value morphisms and on
% the computation morphisms of $\mathcal{M}_1$.
% In particular, we need for every $A$ and $A'$ a reflexive, symmetric relation 
% $\bisim_{A,A'}$ on $\vf(A, A')$ and for every $B$ and $B'$ a reflexive, symmetric relation
% $\bisim_{B,B'}$ on $\ef(B, B')$. Moreover, the relation must respect composition of morphisms.
% This phase of the construction is quite simple, because it consists entirely of extra structure
% supplied as input to the construction.

@@

Suppose we have a \hyperref[def:step-1-model]{step-1 intensional model} $\mathcal{M}$.
Recall that a step-1 intensional model consists of a step-0 model (i.e., a
category internal to the category of CBPV models), along with the necessary
categories and functors for bisimilarity as discussed in Section
\ref{sec:abstract-model-bisimilarity}.
Further, recall that a \hyperref[def:step-2-model]{step-2 model} has everything
a step-1 model has, with the addition of perturbation monoids $\pv_A$ for all
$A$ and $\pe_B$ for all $B$.
% These perturbations must be interpretable as endomorphisms via homomorphisms
% $\ptb_A$ and $\ptb_B$, 
These perturbations must be interpretable via homomorphisms
$\ptb_A$ and $\ptb_B$ as endomorphisms bisimilar to the identity,
and moreover, the push-pull property must hold for all
value relations $c$ and all computation relations $d$.

@@

Now suppose we have a step-2 intensional model $\mathcal{M}$.
We will describe how to construct a \hyperref[def:step-3-model]{step-3 intensional model} $\mathcal{M'}$.
We begin with a definition.

\begin{definition}[representation structure]
Let $c : A \rel A'$ be a value relation. A \emph{left-representation structure} $\rho^L_c$ for $c$ consists of
a value morphism $u \in \vf(A, A')$ such that $c$ is quasi-left-representable by $u$ (see Definition \ref{def:quasi-left-representable}).

Likewise, let $d : B \rel B'$. A \emph{right-representation structure} $\rho^R_d$ for $d$ consists of
a computation morphism $d \in \ef(B', B)$ such that $d$ is quasi-right-representable by $d$ (see Definition \ref{def:quasi-right-representable}).
\end{definition}

(Notice that the direction of the morphism is opposite in the definition of
right-representation structure.)

We now describe the construction of a step-3 model $\mathcal M'$.
The objects of $\mathcal M'$ are defined to be the same as the objects of the
step-2 model $\mathcal M$.
The value and computation morphisms in $\mathcal M'$ are the same as those of
$\mathcal M$.
A value relation is defined to be a pair $(c, \rho^L_c)$ with $c$ a value
relation in $\mathcal M$ and $\rho^L_c$ a left-representation structure for $c$.
Likewise, a computation relation is defined to be a pair $(d, \rho^R_d)$ with
$d$ a computation relation in $\mathcal M$ and $\rho^R_d$ a right-representation
structure for $d$.
Lastly, morphisms of value relations (i.e., value squares) are defined by simply
ignoring the representation structure. That is, a morphism of value relations
$\alpha \in \vsq'((c, \rho^L_c), (c' \rho^L_{c'}))$ is simply a morphism of value
relations in $\vsq(c, c')$. Likewise for computations.

Now we define the functors $F$, $U$, and $\arr$.
On objects, the behavior is the same as the respective functors in $\mathcal M$.
For relations, we define 
$\Fsq' (c, \rho^L_c) = (\Fsq c, \rho^R_{F c})$ and
$\Usq' (d, \rho^R_d) = (\Usq d, \rho^L_{U d})$.
We define $(c, \rho^L_c) \arr (d, \rho^R_d) = (c \arr d, \rho^R_{c \arr d})$.

We now verify that the construction meets the requirements of a step-3 model.
First, we check that composition of value relations (resp. computation relations)
is well-defined. This might seem obvious, but there is actually a subtlety here:
what should the representation structure be for the composition?

>>>>

Lastly, we require that for any value object $A$, the hom-set $\ef(FA, FA)$ contains a
distinguished morphism $\delta_A$, such that $\delta_A \bisim_{FA, FA} \id_{FA}$.

@@

\begin{definition}\label{def:step-1-model}
A \emph{step-1 intensional model} consists of all the data of a step-0 intensional model along
with the additional categories and functors for bisimiarity described above, as  well as
the existence of a distinguised computation morphism $\delta_A \bisim \id_{FA}$ for each $A$.
\end{definition}

@@

We require that $\delta_A \in \pe_{FA}$ for all $A$, where $\delta_A$ is the distinguished
morphism that is required to be present in every hom-set $\ef(FA, FA)$ per the definition
of a step-1 model.

@@

Now we can discuss what it means for an intensional model to model the dynamic type.
This applies to any of the above abstract model definitions, i.e., steps 0-3.

\begin{definition}\label{def:step-4-model}
  A \emph{step-$i$ intensional model with dyn} is a step-$i$ model $\mathcal M$ such that:
  %a distinguished value object $D \in \ob(\vf)$ such that:
  %
  \begin{enumerate}
    \item There is a distinguished value object $D \in \ob(\vf)$.

    \item For each value type $A$, there is a value relation $\text{inj}_A : A \rel D$.

    \item If $c : A \rel A'$, then $\text{inj}_{A} = c \comp \text{inj}_{A'}$.
  \end{enumerate}
\end{definition}

% (By definition of a step-3 model, this relation satisfies the push-pull property and is
% quasi-left-representable.)

% (By definition of a step-3 model, this means there is also a monoid $\pv_D$ of
% perturbations and a homomorphism $\ptbv_D$.)

@@

In the previous section, we have given the definition of an intensional model
of gradual typing as a series of steps with each definition building on the previous one.
%
Here, we discuss how to construct an extensional model from an intensional model with dyn.
We do so in several phases, beginning with a step-1 intensional model with dyn
and ending with an extensional model.
Moreover, this construction is \emph{modular}, in that each phase of the
construction does not depend on the details of the previous ones.
% However, this process cannot proceed in isolation: some phases require
% additional inputs. We will make clear what data must be supplied to each phase.

@@

Suppose we have a \hyperref[def:step-1-model]{step-1 intensional model} $\mathcal{M}$ with dyn.
Recall that a step-1 intensional model consists of a step-0 model (i.e., a
category internal to the category of CBPV models), along with the necessary
categories and functors for bisimilarity as discussed in Section
\ref{sec:abstract-model-bisimilarity}.
Further, recall that a \hyperref[def:step-2-model]{step-2 model} has everything
a step-1 model has, with the addition of perturbation monoids $\pv_A$ for all
$A$ and $\pe_B$ for all $B$.
These perturbations must be interpretable via homomorphisms
$\ptb_A$ and $\ptb_B$ as endomorphisms bisimilar to the identity,
and moreover, the push-pull property must hold for all
value relations $c$ and all computation relations $d$.

@@

Now suppose we have a step-2 intensional model $\mathcal{M}$.
We will describe how to construct a \hyperref[def:step-3-model]{step-3 intensional model} $\mathcal{M'}$.

We now describe the construction of a step-3 model $\mathcal M'$.
